Supplemental data to ``\emph{k}-mer-based GWAS enhances causal variant and candidate gene discovery in soybean''

### 2023-02-28 additional_files/main_text.tex
Adjust the contents of the main and supplementary results to reduce redundancy ; figure order has yet to be adjusted accordingly

@@

% Pubescence color - T locus
\cite{zabala2003} identified Glyma.06g202300 as the gene associated with the
\textit{T} locus for pubescence color and found a 1-bp deletion as
a putative causal variant at this locus. For all three GWAS approaches, we
found a signal overlapping this gene (Figure
\ref{pubescence_color_all_manhattan}). In
particular, the regions defined by the top 1\% most associated variants identified by the
Platypus and \textit{k}-mer approaches included Glyma.06g202300 (Figure
\ref{pubescence_color_all_T_signal}). While the most significant
\textit{k}-mers did occur within the sequence of this gene, they did not
correspond to the causal variant (Figure \ref{pubescence_color_all_T_gene}).
Still, the fourth and fifth most strongly associated \textit{k}-mers
corresponded to the causal variant (Figures \ref{pubescence_color_all_T_gene},
\ref{pubescence_color_all_T_kmers}). As for Platypus, the most significant
variant within the sequence of Glyma.06g202300 was the causal
variant, however it was not the most significantly associated variant overall
(Figures \ref{pubescence_color_all_T_signal},
\ref{pubescence_color_all_T_gene}). Paragraph did not detect the causal
variant at this locus given that this approach does not target small indels.

@@

We conducted a GWAS analysis that only compared accessions with black and brown hila
in an attempt to detect stronger signals at the \textit{R} locus.  We
accordingly detected a signal overlapping this locus using all approaches,
but signals overlapped the Glyma.09g235100 gene associated with the
\textit{R} locus only for Platypus and \textit{k}-mers (Figures
\ref{hilum_color_blackbrown_manhattan}, \ref{hilum_color_blackbrown_R_signal}).
We observed two causal variants previously documented by \cite{gillman2011},
but these were not among the most significantly associated variants or
\textit{k}-mers (Figures \ref{hilum_color_blackbrown_R_gene},
\ref{hilum_color_blackbrown_R_kmers}).  This failure to identify causal
variants more directly could be due to the existence of more than one causal
variant at this locus. Interestingly, the most significantly associated
\textit{k}-mers mapped just upstream of a gene (Glyma.09g234900) identified as
a homolog of Glyma.09g235100 \citep{gillman2011}.

@@

We conducted a GWAS analysis comparing accessions with normal and semi-sparse
pubes\-cence in order to detect loci associated with pubescence density. This analysis
detected signals at the \textit{Ps} locus using all three approaches (Figures
\ref{pubescence-density-main-figure}a-b, \ref{pubescence_density_manhattan}). The causal
variant at this recently cloned locus is a copy number variant overlapping the
Glyma.12g187200 gene, with higher copy number resulting in reduced pubescence \citep{liu2020ps}.
We identified a region of $\sim$42 kb that varied in copy number across
the population and was associated with the trait (Figure
\ref{pubescence-density-main-figure}c), and therefore possibly corresponded to the
25.6-kb region identified as the causal CNV by \cite{liu2020ps}. We found that
reads containing the most significantly associated \textit{k}-mer mapped to either of the two ends
of the CNV (Figure \ref{pubescence-density-main-figure}b,d). This suggests that
the most associated \textit{k}-mer is indeed linked to the CNV
that is causal at this locus, while \textit{k}-mers within the
CNV region may represent differences in sequence between the copies.

@@

Next, we conducted a GWAS analysis comparing dull and shiny seed coat phenotypes
for the seed coat luster trait. Seed coat luster has been shown by \cite{gijzen1999}
to be largely caused by the deposition of a hydrophobic protein (HPS)
at the seed surface. Copy number variation of this gene at the \textit{B}
locus was later shown to explain variation in luster, although sequence
variation in the HPS sequence may also play a role \citep{gijzen2006}.
We found signals overlapping the causal CNV using all approaches, and the
most significantly associated \textit{k}-mer was notably located
within the CNV region (Figures \ref{seed_coat_luster_dullshiny_manhattan},
\ref{seed_coat_luster_dullshiny_B_signal}). However, we were not able
to directly link any of the variants or \textit{k}-mers identified to the causal
CNV.

% Maturity group loci
\subsubsection*{Maturity}
\label{sv-gwas-main-results-maturity}

We did not detect any signals associated with maturity using our dataset
(Figure \ref{maturity_group_manhattan}). The use of maturity group as a proxy
for time to maturity instead of the number of days, combined with the limited
phenotypic variation of this trait in this collection of accessions (Table
\ref{MATURITY.GROUP_gwas_table}), likely explains our failure to detect
any signals.

% Seed oil and protein
\subsubsection*{Seed oil and protein content}
\label{sv-gwas-main-results-oil-protein}

We expected to detect signals at two loci controlling seed oil and protein
content on chromosomes 15 and 20 \citep{bandillo2015}. However, none of our
analyses detected signals overlapping the genes cloned for these loci (Figures
\ref{oil_manhattan}, \ref{protein_manhattan}). This is probably due to the fact
that the oil and protein content data that we used for GWAS were not obtained
from orthogonal trials, which was a clear limitation to this GWAS given our
relatively small sample size.

@@

The two known loci for pubescence form in soybean, \textit{Pa1} and
\textit{Pa2}, have also not been cloned yet, although \cite{gilbert2017} has suggested
Glyma.12g213900 as a candidate for \textit{Pa1} based on data published
by \cite{bandillo2017}. This gene encodes a Myb transcription factor annotated
as playing a role in trichome branching. Indeed, the most
significant signals found by all three approaches were located near this gene.
analyses (Figures \ref{pubescence-form-main-figure}a-b,
\ref{pubescence_form_all_manhattan}, \ref{pubescence_form_all_Pa1_signal}).
The analyses based on Platypus and \emph{k}-mers identified two highly
significant non-synonymous SNPs in Glyma.12g213900 that were associated with appressed pubescence
(Figures \ref{pubescence-form-main-figure}c-d, \ref{pubescence_form_all_Pa1_gene},
\ref{pubescence_form_all_Pa1_kmers}). The alternative alleles for these two
SNPs co-occurred together in our dataset and may represent causal variants at
the \emph{Pa1} locus. The signals observed at the \emph{Pa2} locus were much
weaker and the signal observed from the \emph{k}-mer GWAS overlapped a total of
20 genes (Figures \ref{pubescence_form_all_manhattan},
\ref{pubescence_form_all_Pa2_signal}), none of which appeared obviously linked
to pubescence form.

@@

In addition to verifying known loci, we also observed potentially novel loci in
the \textit{k}-mer GWAS analysis of stem termination and pod color.
When comparing accessions
with indeterminate and semi-determinate stem termination types, we observed
previously undocumented signals on chromosomes Gm11, Gm16 and Gm18 in
addition to the known \textit{Dt1} locus (Figure \ref{stem-termination-main-figure}a).
Analysis of the pairwise LD between significant \textit{k}-mers found no
LD between these signals, suggesting that they may represent \textit{bona fide}
loci associated with stem termination type
(Figure \ref{stem-termination-main-figure}b). Similarly, a \textit{k}-mer GWAS
analysis comparing accessions with black and brown pods found a previously
undocumented signal on chromosome Gm15. \textit{k}-mers at this new locus showed no LD with \textit{k}-mers
at the known \textit{L1} locus and may therefore represent a distinct locus (Figures
\ref{pod_color_blbr_manhattan}, \ref{pod_color_blbr_ld}).

>>>>

The \textit{k}-mer GWAS analysis also identified presumably spurious significant
\textit{k}-mers scattered throughout the genome (Figure
\ref{flower_color_manhattan}). An analysis of the pairwise LD between a subset
of those \textit{k}-mers showed that most \textit{k}-mers formed a single LD block
linked to the \textit{W1} locus (Figure \ref{flower_color_ld}). A small LD block
occurred on chromosome Gm09, however it was in moderate LD with the one on chromosome
Gm13 and therefore probably does not represent a separate locus.

@@

% Pubescence color - T locus
\cite{zabala2003} identified Glyma.06g202300 as the gene associated with the
\textit{T} locus for pubescence color and found a 1-bp deletion as
a putative causal variant at this locus. For all three GWAS approaches, we
found a signal overlapping this gene (Figure
\ref{pubescence_color_all_manhattan}). In
particular, the regions defined by the top 1\% most associated variants identified by the
Platypus and \textit{k}-mer approaches included Glyma.06g202300 (Figure
\ref{pubescence_color_all_T_signal}). While the most significant
\textit{k}-mers did occur within the sequence of this gene, they did not
correspond to the causal variant (Figure \ref{pubescence_color_all_T_gene}).
Still, the fourth and fifth most strongly associated \textit{k}-mers
corresponded to the causal variant (Figures \ref{pubescence_color_all_T_gene},
\ref{pubescence_color_all_T_kmers}). We
observed that not all accessions with gray pubescence harbored the deleted
nucleotide (Figure \ref{pubescence_color_all_T_kmers}c), which could indicate either that other causal variants are resulting in
a non-functional F3’H in this population, or that the causal
variant lies elsewhere. As for Platypus, the most significant
variant within the sequence of Glyma.06g202300 was the causal
variant, however it was not the most significantly associated variant overall
(Figures \ref{pubescence_color_all_T_signal},
\ref{pubescence_color_all_T_gene}). Paragraph did not detect the causal
variant at this locus given that this approach does not target small indels.

@@

We conducted a GWAS analysis that only compared accessions with black and brown
hila in an attempt to detect stronger signals at the \textit{R} locus.  We
accordingly detected a signal overlapping this locus using all approaches, but
only Platypus and \textit{k}-mers detected signals overlapping the
Glyma.09g235100 gene associated with \textit{R} (Figures
\ref{hilum_color_blackbrown_manhattan}, \ref{hilum_color_blackbrown_R_signal}).
\cite{gillman2011} identified this R2R3 MYB transcription factor as the
molecular basis for the \textit{R} locus and documented four different
loss-of-function alleles linked to brown hilum color. Although our analyses did
detect strong signals in the vicinity of this gene, the most associated markers
were not located in the body of the gene (Figure
\ref{hilum_color_blackbrown_R_signal}).  Interestingly, the most significantly
associated \emph{k}-mers detected by our analysis mapped 2 kb upstream of
another gene (Glyma.09g234900) annotated as a MYB transcription factor and
putatively involved in the regulation of anthocyanin biosynthesis. This gene as
well as Glyma.09g235100 and two others are part of a tandem array of four R2R3
MYB genes considered as candidates by \cite{gillman2011}, and of which
Glyma.09g235100 was the only one expressed in the seed coat. Given the
convincing evidence provided by \cite{gillman2011}, it is most likely that
Glyma.09g235100 is indeed the gene associated with the \emph{R} locus and that
our failure to identify the most significantly associated \emph{k}-mers as
causal variants is due to the coexistence of several variants that can cause
loss of function at the \emph{R} locus.  Still, we observed two causal variants
previously documented by \cite{gillman2011} in our dataset (Figures
\ref{hilum_color_blackbrown_R_gene}, \ref{hilum_color_blackbrown_R_kmers}).

@@

We conducted a GWAS analysis comparing accessions with normal and semi-sparse
pubes\-cence in order to detect loci associated with pubescence density. This analysis
detected signals at the \textit{Ps} locus using all three approaches (Figures
\ref{pubescence-density-main-figure}a-b, \ref{pubescence_density_manhattan}, \ref{pubescence_density_Ps_signal}
and \ref{pubescence_density_Ps_gene}). The causal
variant at this recently cloned locus is a copy number variant overlapping the
Glyma.12g187200 gene, with higher copy number resulting in reduced pubescence \citep{liu2020ps}.
We identified a region of $\sim$42 kb that varied in copy number across
the population and was associated with the trait (Figure
\ref{pubescence-density-main-figure}c), and therefore possibly corresponded to the
25.6-kb region identified as the causal CNV by \cite{liu2020ps}. We found that
reads containing the most significantly associated \textit{k}-mer mapped to either of the two ends
of the CNV (Figure \ref{pubescence-density-main-figure}b,d). This suggests that
the most associated \textit{k}-mer is indeed linked to the CNV
that is causal at this locus, while \textit{k}-mers within the
CNV region may represent differences in sequence between the copies.

@@

Next, we conducted a GWAS analysis comparing dull and shiny seed coat
phenotypes for the seed coat luster trait. Seed coat luster has been shown by
\cite{gijzen1999} to be largely caused by the deposition of a hydrophobic
protein (HPS) at the seed surface. Copy number variation of this gene at the
\textit{B} locus was later shown to explain variation in luster, although
sequence variation in the HPS sequence may also play a role \citep{gijzen2006}.
Using a sequencing depth-based analysis, we identified a 31-kb region spanning
positions 9,386,109-9,417,431 on chromosome Gm15 that exhibited variation in
copy number across accessions. This CNV overlapped four genes annotated as
containing HPS domains and therefore likely represented the causal variant
(Figures \ref{seed_coat_luster_dullshiny_B_signal},
\ref{seed_coat_luster_dullshiny_B_gene}).  We found signals overlapping this
CNV using all approaches, and the most significantly associated \textit{k}-mer
was notably located within the CNV region (Figures
\ref{seed_coat_luster_dullshiny_manhattan},
\ref{seed_coat_luster_dullshiny_B_signal}). However, we were not able to
directly link any of the variants or \textit{k}-mers identified to the causal
CNV.

@@

The two known loci for pubescence form in soybean, \textit{Pa1} and
\textit{Pa2}, have also not been cloned yet, although \cite{gilbert2017} has suggested
Glyma.12g213900 as a candidate for \textit{Pa1} based on data published
by \cite{bandillo2017}. This gene encodes a Myb transcription factor annotated
as playing a role in trichome branching. Indeed, the most
significant signals found by all three approaches were located near this gene
(Figures \ref{pubescence-form-main-figure}a-b,
\ref{pubescence_form_all_manhattan}, \ref{pubescence_form_all_Pa1_signal}).
The analyses based on Platypus and \emph{k}-mers identified two highly
significant non-synonymous SNPs in Glyma.12g213900 that were associated with appressed pubescence
(Figures \ref{pubescence-form-main-figure}c-d, \ref{pubescence_form_all_Pa1_gene},
\ref{pubescence_form_all_Pa1_kmers}). The alternative alleles for these two
SNPs co-occurred together in our dataset and may represent causal variants at
the \emph{Pa1} locus. The signals observed at the \emph{Pa2} locus were much
weaker and the signal observed from the \emph{k}-mer GWAS overlapped a total of
20 genes (Figures \ref{pubescence_form_all_manhattan},
\ref{pubescence_form_all_Pa2_signal}), none of which appeared obviously linked
to pubescence form.

A second GWAS analysis focused only on appressed and semi-appressed phenotypes
with the objective of enhancing the signal at the \emph{Pa2} locus. However,
this analysis only detected weak signals at the \emph{Pa1} locus (Figure
\ref{pubescence_form_noerect_manhattan}) and its results were therefore not
analyzed further.

@@

In addition to verifying known loci, we also observed potentially novel loci in
the \textit{k}-mer GWAS analysis of stem termination and pod color.
When comparing accessions
with indeterminate and semi-determinate stem termination types, we observed
previously undocumented signals on chromosomes Gm11, Gm16 and Gm18 in
addition to the known \textit{Dt1} locus (Figure \ref{stem-termination-main-figure}a).
Analysis of the pairwise LD between significant \textit{k}-mers found no
LD between these signals, suggesting that they may represent \textit{bona fide}
loci associated with stem termination type
(Figure \ref{stem-termination-main-figure}b). Similarly, a \textit{k}-mer GWAS
analysis comparing accessions with black and brown pods found a previously
undocumented signal on chromosome Gm15. \textit{k}-mers at this new locus showed no LD with \textit{k}-mers
at the known \textit{L1} locus and may therefore represent a distinct locus (Figures
\ref{pod_color_blbr_manhattan}, \ref{pod_color_blbr_ld}). The observed signal overlaps a total of
27 genes (Figure \ref{pod_color_blbr_pdcGm15_signal}), but none of those are
annotated as having an obvious role in controlling pod color.
